Disclosing funding sources for open access publication fees: the Open APC

### 2016-05-29 article.Rmd
new draft discussion

The Open APC initiative extends existing methods to disclose spending on open access publication fees. Our workflow benefits from openly available tools and the social coding platforms GitHub, both of which are well established and suited to increase transparency in research [@Peng_2011; @Ram_2013]. For 99 % of the articles, CrossRef provided bibliographic information, which substantially contributed to a uniform data-set about formerly fragmented payments made for open access articles.

Although CrossRef disambiguates journal titles and publisher names and is therefore an authority-controlled source for open access journal literature, derivations from CrossRef metadata curation as well as the context of aggregation must be made clear. In particular, problems persist on how to deal with name changes and ongoing mergers. For example, the publisher Public Library of Science (PLOS) has changed its acronym from "PLoS" to "PLOS", which CrossRef metadata reflects from 2015 onward. We therefore normalized all PLOS journal titles in order to secure unique reference to these journals. Another publisher affected is "The Optical Society," formerly "Optical Society of America". Because the ownership of publishing houses can become combined, dealing with mergers is also essential to make cost data comparable. Jisc data, for instance, differentiate between the full open access publisher BioMed Central and the traditional publisher Springer, concluding that "traditional publishing houses" lean on the hybrid model [@Pinfield_2015]. This stands in stark contrast to our approach, in which CrossRef metadata reflects the merger of BioMed Central and Springer, resulting in Springer Science + Business Media to be the best represented publisher for articles in full open access journals in the Open APC data-set. Another approach for ensuring unique reference, but that we have not evaluated yet, is to use CrossRef's identifiers for journals and publishers instead.[^19]

Because of the dynamic landscape of academic publishing and its representation in CrossRef's data curation efforts, it is important to consider the time-frame of metadata aggregation. In our case, we re-used metadata shortly after the data submission. However, for some cost analysis -- for instance to prepare negotiations with publishers on future schemes to fund open access journal articles -- it could be more feasible to re-normalize the complete data. The rossRef API provides incremental metadata updates that can be used to assess the current potential of future funding. While licensing information is incompletely covered in the CrossRef index so far, and therefore not analyzed in our study, the growing importance of facilitating text mining may result in more and more publishers sharing this information with CrossRef in the future.

Participation is voluntary. Therefore, not all institutions in Germany that provide central funding of publication fees contribute cost data to this initiative. In a qualitative survey that also asked why German institutions are reluctant to share their cost data through the Open APC initiative one institution feared that increase in transparency would allow publishers to adjust prices in their favor. Others pointed out that the workload to produce such a data-set could be too extensive [@deppe_2015].

While there may still be institutions that have no overview of their APC spending, we would like to emphasize that reporting data that is already available within an institution to the Open APC initiative should not lead to much additional work. The central incentive of this initiative is to make it as easy as possible to submit data to it. Being able to combine that data into one standardized data-set increases transparency and comparability, and gives institutions a better understanding of the overall development of open access publishing. In particular, we cannot see the harm of increased transparency; in fact, not knowing how much is spent is undoubtedly a disadvantage in dealing with publishers.

Extending the data template to include information about funders or whether special agreements with publishers applied as suggested by @Pinfield_2015, would even increase the efforts needed to participate. However, with the growing demand for action in areas like the large-scale transition of toll-access journals to open access [@Schimmer_Geschuhn_Vogler], an updated data template could help institutions to better comply with these policy developments in the future. From our experience, another barrier to participate is the lack of skills in version control: the data submission itself is not always made directly by the institutions. Instead, they sent files to Bielefeld University Library with the request to make them available on GitHub on their behalf.

Future work needs to focus on analyzing the cost data. Of particular interest are questions concerning the coverage of central funding schemes in comparison with open access publication output in general, and the use of other means to cover publication fees in particular. The Open APC initiative does not cover personal budgets or make price reductions explicit, but studies suggest that there is a possible gray area [@BJ_RK_2012; @Bj_rk_2015; @Lawson_waiver_2015]. Existing study designs could be re-applied to examine the relationship between price on the one hand, and indexing coverage, journal prestige or management costs on the other [@Walters_2011; @Bj_rk_scientometrics_2015, @Pinfield_2015]. This, in turn, helps to address the central question of future business models in scholarly publishing from an international perspective [@Pinfield_2015].

<!-- # Conclusion{.unnumbered}

The purpose of the Open APC initiative is to gain better insight on spending for open access. The main operational idea—to make data submission easy by building upon existing data and by using popular open tools and data bases—has proven to be a feasible approach and to scale well. Starting with a few institutions, the integration of further data contributors has not been a problem. While we use article metadata reported by publishers to CrossRef, we think it's best to collect information on how much has been spent at the sources: the institutions paying for APCs. We look forward to more data contributors to the initiative and to adapt this scheme to new analyses, additional data, or international sources. The Open APC initiative can serve as an example of the value that open data, open source software, and the knowledge and skills of tech-savvy librarians add to modern tasks of libraries.-->

>>>>

In Germany, institutional spending on open access publication fees has increased over the years. With a share of …, the majority of open access articles German institutions reported to the Open APC initiative were published in fully open access journals. This presumably reflects the DFG funding policy which excludes the support of articles published in hybrid open access journals. The DFG has been financially supporting the implementation of central publication funds at more than 30 German universities since 2009. However, reviewing self-reported cost data from funders or countries that also support hybrid open access journals or open access books revealed a smaller proportions of payments in favor of articles in fully open access journals. Since open access publication fee spending is generally fragmented, we cannot answer whether German researchers avoid opting for open access when publishing in hybrid journals or use other budgets to pay publication fees required to make their work open access through these kind of journals.

In our study, CrossRef thoroughly indexed the analysed open access journal articles. Metadata representing publisher and journal titles could be gathered for 99 % of the self-reported article and successfully merged with the Open APC cost data. Using metadata from CrossRef, therefore, could reduce the extensive validation work of bibliographic information in self-reported data-sets on open access publication fees provided that the reporting of the DOI along with the expenses is made mandatory. Drawing on CrossRef would also increase the comparability of cost data for future negotiations with publishers on open access agreements, and the open access spending between open data initiatives that apply the same reporting standards, as its metadata represent the dynamic landscape of academic publishing in terms of ongoing mergers of publishing houses or name changes.

Another advantage of self-reported data-sets on the article-level is that they enable researchers and practitioners alike to study in which open access journals researchers from one institution actually publish, and to compare these findings with that of other universities or research organizations. For instance, our study revealed that the size of publication fee spending differed among the institutions with the  Max Planck Society accounting for almost 39 % of the overall articles. Many universities and research organization reported remarkably lower number of supported open access articles to the Open APC initiative. Using self-reported data, therefore, contributes to the understanding about how much and to what extent spending on open access publishing varies on the institutional level. This is particularly relevant given the increasingly important role open access publishing plays in recent negotiations between German universities and research organizations forming consortia on the one side and publishers on the other about financing scholarly publishing in future [allianz].

Our study is limited in some respects. One is that we cannot assess whether publishers and journals granted publication fee discounts seeing that the Open APC initiative does not track this kind of information. However, the large price ranges of particular journals suggests that varying pricing schemes are in place. Adding to this complexity, it seems that some institutions only paid parts of the publication fee. Take for instance the journal Nature Communication. Charges reported in our sample ranged between 2000 €, the DFG price cap, and 4.403 €. Such co-payments that involve several budgets are a discussed strategy to sustain publication funds at German universities [Fournier]. In another case, one university included its charges for participating in the German SCOAP consortia and presumably divided the sum by the articles published in SCOAP journals. In this consortia managed by the German National Library of Science and Technology payments were not directly made on the article level. Instead, subscription costs between a participating library and a publisher were reconciled, and the reduction transferred to the consortia to finance publications in SCOAP journals.

It must also be noted that reporting to the Open APC initiative is voluntary. Therefore, not all institutions in Germany that provide central funding of publication fees contribute cost data to this initiative. In a qualitative survey that also asked why German institutions are reluctant to share their cost data through the Open APC initiative one institution feared that increase in transparency would allow publishers to adjust prices in their favor. Others pointed out that the workload to produce such a data-set could be too extensive [@deppe_2015]. As there no registry of institutional open access funds or similar support structures exists, we cannot assess how many German universities and research organizations do not share their spendings on open access publication fees.

Our analysis on how institutional spending per articles was distributed over publishers and journal titles shows that open access publishing is diverse and concentrated at the same time. While we were able to identify 139 individual publishing houses that were supported by the German universities and research organizations, the distribution is highly skewed. 92 % of open access publication fee spending went to ten publishers, confirming a general high concentration of few publishers in current academic publishing. However, our study could not confirm that publications in open access journals owned by traditional publishing houses account for most of the spending on publication fees. Rather, open access publishers such as Public Library of Science (PloS), Copernicus GmbH or MPDI AG rank higher in our study than in the analyses of cost data in the UK. The study finally confirms the importance of mega-journals that account for a large share of all articles in fully open access journals. 
Conclusion

# Conclusion{.unnumbered}

<!-- The purpose of the Open APC initiative is to gain better insight on spending for open access. The main operational idea—to make data submission easy by building upon existing data and by using popular open tools and data bases—has proven to be a feasible approach and to scale well. Starting with a few institutions, the integration of further data contributors has not been a problem. While we use article metadata reported by publishers to CrossRef, we think it's best to collect information on how much has been spent at the sources: the institutions paying for APCs. We look forward to more data contributors to the initiative and to adapt this scheme to new analyses, additional data, or international sources. The Open APC initiative can serve as an example of the value that open data, open source software, and the knowledge and skills of tech-savvy librarians add to modern tasks of libraries.-->
